---
title: "Multimodal data" 
author: "BS. Mauricio Morales"
date: "`r Sys.Date()`"
format:
    html:
        theme:
            dark: darkly
output: html_document
editor:
    markdown:
        wrap: 60
---

# Install and load packgages

```{r}
#| warning: false

packages <- c(
    "renv",
    "dplyr",
    "purrr",
    "here",
    "tidyr",
    "MetBrewer",
    "ggplot2",
    "patchwork",
    "magrittr",
    "ComplexHeatmap",
    "devtools",
    # Single Cell
    "Seurat",
    "SeuratObject",
    "SeuratData",
    # Single Cell plotting
    "SCpubr",
    # Fast Wilcoxon rank sum test and auROC analysis
    "presto",
    # Meta-analysis
    "metap"
)
# Install
installed_packages <- packages %in%
    rownames(installed.packages())
if (any(installed_packages == F)) {
    install.packages(packages[!installed_packages])
}
# Load
suppressPackageStartupMessages({
    invisible(lapply(packages, library, character.only = T))
})
```

# Create project-local library

```{r}
#| eval: false

# If you add new libraries perform an snapshot
renv::snapshot()
# Open configuration files
usethis::edit_r_environ(scope = "user")
# Source global helper functions
source(file = file.path(directory, "src/code_base.R"))
```

# Set working directory

```{r}
directory <- here::here()
```

# Load and set up dataset

```{r}
#| eval: false

# Install dataset
SeuratData::InstallData("ifnb")

# Load
ifnb <- SeuratData::LoadData("ifnb")

# Split the RNA measurements into two layers one for control cells, one for stimulated cells
ifnb[["RNA"]] %<>% 
    split(f = ifnb$stim)

# Save
# saveRDS(
#     object = ifnb,
#     file = file.path(
#         directory,
#         "outputs",
#         "02_integration",
#         paste0(format(Sys.Date(), "%b_%d_%Y"),
#         "_ifnb_raw.rds")
#     )
# )
```

# Perform analysis without integration

```{r}
# Create outputs directories if do not exist yet
fs::dir_create(
    path = file.path(
        directory,
        c("plots", "outputs"),
        "02_integration"
    )
)

# Perform analysis
ifnb %<>%
    Seurat::NormalizeData() %>%
    Seurat::FindVariableFeatures() %>%
    Seurat::ScaleData() %>%
    Seurat::RunPCA() %>%
    Seurat::FindNeighbors(
        dims = 1:30, 
        reduction = "pca") %>%
    Seurat::FindClusters(
        resolution = 2) %>%
    Seurat::RunUMAP(
        dims = 1:30, 
        reduction = "pca", 
        reduction.name = "umap.unintegrated")

# Save
# saveRDS(
#     object = ifnb,
#     file = file.path(
#         directory,
#         "outputs",
#         "02_integration",
#         paste0(format(Sys.Date(), "%b_%d_%Y"),
#         "_ifnb_processed.rds")
#     )
# )

# Call helper function for UMAP
.stim_umap <- .dim_umap_scpub(
    SEURAT_OBJECT = ifnb,
    POINT_SIZE = 1,
    LABEL = FALSE,
    REDUCTION = "umap.unintegrated",
    GROUP_BY = "stim",
    PALETTE = "Archambault"
) +
    ggplot2::ggtitle(
        "Groups"
    )

.clusters_umap <- .dim_umap_scpub(
    SEURAT_OBJECT = ifnb,
    POINT_SIZE = 1,
    LABEL = FALSE,
    REDUCTION = "umap.unintegrated",
    GROUP_BY = "seurat_clusters",
    PALETTE = "Signac"
) + 
    ggplot2::ggtitle(
        "Clusters"
    )

# Save
# .tiff_pdf_plot(
#     PLOT = .stim_umap + .clusters_umap,
#     HEIGHT = 6,
#     WIDTH = 8,
#     DIRECTORY = file.path(
#         directory,
#         "plots",
#         "02_integration"),
#         FILENAME = paste0(
#             format(Sys.Date(), "%b_%d_%Y"),
#         "_umap_stim_plus_clusters"))
```

# Perform analysis with integration

TODO: helper function from code base, generated a misleading umap when using `SCpubr`, **after integration**, check what is the source of this issue. 

DONE: wrong reduction layers for plotting, correction: `integrated.cca` -> `umap` (in plotting)

```{r}
ifnb.integrated <- Seurat::IntegrateLayers(
    object = ifnb,
    method = CCAIntegration,
    orig.reduction = "pca",
    new.reduction = "integrated.cca",
    verbose = FALSE
    ) %>%
    SeuratObject::JoinLayers(
        assay = "RNA"
    ) %>%
    Seurat::FindNeighbors(
        reduction = "integrated.cca",
        dims = 1:30
    ) %>%
    Seurat::FindClusters(
        resolution = 1
    ) %>%
    Seurat::RunUMAP(
        dims = 1:30, 
        reduction = "integrated.cca")

# Save
# saveRDS(
#     object = ifnb.integrated,
#     file = file.path(
#         directory,
#         "outputs",
#         "02_integration",
#         paste0(format(Sys.Date(), "%b_%d_%Y"),
#         "_ifnb_layers_integrated.rds")
#     )
# )

# UMAP post layer integration
.stim_umap <- .dim_umap_scpub(
    SEURAT_OBJECT = ifnb.integrated,
    POINT_SIZE = 1,
    LABEL = FALSE,
    REDUCTION = "umap",
    GROUP_BY = "stim",
    PALETTE = "Archambault") +
    ggplot2::ggtitle(
        "Groups")

.clusters_umap <- .dim_umap_scpub(
    SEURAT_OBJECT = ifnb.integrated,
    POINT_SIZE = 1,
    LABEL = FALSE,
    REDUCTION = "umap",
    GROUP_BY = "seurat_annotations",
    PALETTE = "Signac") + 
    ggplot2::ggtitle(
        "Clusters")

# Save
# .tiff_pdf_plot(
#     PLOT = .stim_umap + .clusters_umap,
#     HEIGHT = 6,
#     WIDTH = 8,
#     DIRECTORY = file.path(
#         directory,
#         "plots",
#         "02_integration"),
#         FILENAME = paste0(
#             format(Sys.Date(), "%b_%d_%Y"),
#         "_umap_stim_plus_clusters_layers_integrated"))

# UMAP sply by facet based on stim
.clusters_umap_split <- .dim_umap_scpub(
    SEURAT_OBJECT = ifnb.integrated,
    POINT_SIZE = 1,
    LABEL = FALSE,
    REDUCTION = "umap",
    SPLIT_BY = "seurat_clusters",
    GROUP_BY = "stim",
    PALETTE = "Archambault") + 
    ggplot2::ggtitle(
        "Clusters")

# Save
# .tiff_pdf_plot(
#     PLOT = .clusters_umap_split,
#     HEIGHT = 12,
#     WIDTH = 12,
#     DIRECTORY = file.path(
#         directory,
#         "plots",
#         "02_integration"),
#         FILENAME = paste0(
#             format(Sys.Date(), "%b_%d_%Y"),
#         "_umap_stim_split_layers_integrated"
#         ))
```

# Identify conserved markers

```{r}
# NK example
NK_MARKERS <- ifnb.integrated %>%
    Seurat::SetIdent(value = "seurat_annotations") %>%
    Seurat::FindConservedMarkers(
        ident.1 = "NK",
        grouping.var = "stim",
        verbose = FALSE
    )


```

```{r}
sessionInfo()
```