---
title: "Dictionary Learning for cross-modality integration" 
author: "BS. Mauricio Morales"
date: "`r Sys.Date()`"
format:
    html:
        theme:
            dark: darkly
output: html_document
editor:
    markdown:
        wrap: 60
---

# Install and load packgages

```{r}
#| warning: false

packages <- c(
    "renv",
    "dplyr",
    "purrr",
    "here",
    "tidyr",
    "MetBrewer",
    "ggplot2",
    "patchwork",
    "magrittr",
    "ComplexHeatmap",
    "devtools",
    "magrittr",
    # Single Cell
    "Seurat",
    "SeuratObject",
    "SeuratData",
    # Single Cell plotting
    "SCpubr",
    # Single-cell chromatin data
    "Signac",
    # Genome related
    "GenomeInfoDb",
    "EnsDb.Hsapiens.v86",
    # Python alternatives
    "reticulate"
)
# Install
installed_packages <- packages %in%
    rownames(installed.packages())
if (any(installed_packages == F)) {
    install.packages(packages[!installed_packages])
}
# Load
suppressPackageStartupMessages({
    invisible(lapply(packages, library, character.only = T))
})
```

# Create project-local library

```{r}
#| eval: false

# If you add new libraries perform an snapshot
renv::snapshot()

# Open configuration files
usethis::edit_r_environ(scope = "user")

# Source global helper functions
source(file = file.path(directory, "src/code_base.R"))
```

# Set working directory

```{r}
directory <- here::here()

# Allocate 10GB of memory
options(future.globals.maxSize = 10 * 1024^3)
```

# Set uv enviroment

Note 1: due to incompatibility with MacOS and `hdf5r`, alternative libraries will be use. `Python` is suitable for this: `reticulate`. 

Note 2: `uv` for virtual env management was used:
- Activate via Positron.
- Installed via brew in bash-terminal.
- Sync same version present in Positron via `❯ uv venv ~/.virtualenvs/r-reticulate-311 --python 3.13`
- Install libraries before calling `py_require`: `uv pip install --python ~/.virtualenvs/r-reticulate-313/bin/python "numpy<2.5" scanpy`

```{r}
# Indicate virtual env path (using uv)
reticulate::use_virtualenv("~/.virtualenvs/r-reticulate-313", required = TRUE)
```

# Load the bridge, query, and reference datasets

## Load and setup the 10x multiome object

Note 3: to proceed with the tutorial is necessary to workaround with solutions to avoid `hdf5r` library or related dependencies. Strategy: use `scanpy` to read files and proceed with proper conversion before start working with `seurat`.

```{r}
# Initialize library (`convert = FALSE` prevent for `Python`->`R` conversion)
sc <- reticulate::import("scanpy", convert = FALSE)

# Read file
input10x_h5 <- sc$read_10x_h5(
    file.path(
        directory,
        "inputs",
        "cross_modality_integration",
        "pbmc_granulocyte_sorted_10k_filtered_feature_bc_matrix.h5"
    ),
    gex_only = FALSE
)

# Remove duplicates
input10x_h5$var_names_make_unique()

# Extract features
feature_types <- reticulate::py_to_r(input10x_h5$var$feature_types$tolist())
rna_idx <- feature_types == "Gene Expression"
atac_idx <- feature_types == "Peaks"

# Get cell and feature names
cell_names <- reticulate::py_to_r(input10x_h5$obs_names$tolist())
feature_names <- reticulate::py_to_r(input10x_h5$var_names$tolist())

# Convert sparse matrix
x_csc <- input10x_h5$X$tocsc()
count_mtx <- reticulate::py_to_r(x_csc)
count_mtx <- Matrix::t(count_mtx)

# Assign column names
colnames(count_mtx) <- cell_names
rownames(count_mtx) <- feature_names

# Split modalities
rna_counts <- count_mtx[rna_idx, ]
atac_counts <- count_mtx[atac_idx, ]

# Create Seurat object with RNA
obj_multi <- Seurat::CreateSeuratObject(counts = rna_counts)

# Create directories
fs::dir_create(
    path = file.path(
        directory,
        c("outputs", "plots"),
        "04_cross_modality_integration"))

# Save
# saveRDS(
#     object = obj_multi,
#     file = file.path(
#         directory,
#         "outputs",
#         "04_cross_modality_integration",
#         paste0(format(Sys.Date(), 
#         "%b_%d_%Y"),
#         "_pbmc_granulocyte_rna_filtered_feature_bc_raw_seurat.rds")))

# Mitochondrial percentage
obj_multi[["percent.mt"]] <- Seurat::PercentageFeatureSet( 
    obj_multi,
    pattern = "^MT-"
)

# ATAC - filter to standard chromosomes
grange_counts <- Signac::StringToGRanges(
    rownames(atac_counts),
    sep = c(":", "-")
)
grange_use <- GenomeInfoDb::seqnames(grange_counts) %in%
    GenomeInfoDb::standardChromosomes(grange_counts)
atac_counts <- atac_counts[as.vector(grange_use), ]

# Gene annotations
annotations <- Signac::GetGRangesFromEnsDb(
    ensdb = EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86
)
GenomeInfoDb::seqlevelsStyle(annotations) <- "UCSC"
GenomeInfoDb::genome(annotations) <- "hg38"

# Create chromatin assay
chrom_assay <- Signac::CreateChromatinAssay(
    counts = atac_counts,
    sep = c(":", "-"),
    genome = "hg38",
    fragments = file.path(
        directory,
        "inputs",
        "cross_modality_integration",
        "pbmc_granulocyte_sorted_10k_atac_fragments.tsv.gz"
    ),
    min.cells = 10,
    annotation = annotations
)

# Save
# saveRDS(
#     object = chrom_assay,
#     file = file.path(
#         directory,
#         "outputs",
#         "04_cross_modality_integration",
#         paste0(format(Sys.Date(), 
#         "%b_%d_%Y"),
#         "_pbmc_granulocyte_multiome_filtered_feature_bc_raw_seurat.rds")))

# Add ATAC assay
obj_multi[["ATAC"]] <- chrom_assay

# QC filter
obj_multi %<>% subset(
    subset = nCount_ATAC < 7e4 &
        nCount_ATAC > 5e3 &
        nCount_RNA < 25000 &
        nCount_RNA > 1000 &
        percent.mt < 20
)

# Save
# saveRDS(
#     object = obj_multi,
#     file = file.path(
#         directory,
#         "outputs",
#         "04_cross_modality_integration",
#         paste0(format(Sys.Date(),
#         "%b_%d_%Y"),
#         "_pbmc_post_qc_rna_multiome.rds")))
```

## Load and setup the 10x scATAC-seq query 

```{r}
# Clean env
rm(list = setdiff(ls(), c("directory", "obj_multi", "input10x_h5")))

# Initialize library (`convert = FALSE` prevents Python->R auto-conversion)
sc <- reticulate::import("scanpy", convert = FALSE)

# Load ATAC dataset
input10x_atac <- sc$read_10x_h5(
    file.path(
        directory,
        "inputs",
        "cross_modality_integration",
        "10k_PBMC_ATAC_nextgem_Chromium_X_filtered_peak_bc_matrix.h5"
    ),
    gex_only = FALSE
)

# Make variable names unique
input10x_atac$var_names_make_unique()

# Extract count matrix
cell_names <- reticulate::py_to_r(input10x_atac$obs_names$tolist())
feature_names <- reticulate::py_to_r(input10x_atac$var_names$tolist())

atac_pbmc_data <- input10x_atac$X$tocsc() |>
    reticulate::py_to_r() |>
    Matrix::t()

colnames(atac_pbmc_data) <- cell_names
rownames(atac_pbmc_data) <- feature_names

# Get gene annotations
annotation <- Signac::GetGRangesFromEnsDb(
    ensdb = EnsDb.Hsapiens.v86::EnsDb.Hsapiens.v86
)
GenomeInfoDb::seqlevelsStyle(annotation) <- "UCSC"

# Create ChromatinAssay for ATAC data
atac_pbmc_assay <- Signac::CreateChromatinAssay(
    counts = atac_pbmc_data,
    sep = c(":", "-"),
    fragments = file.path(
        directory,
        "inputs",
        "cross_modality_integration",
        "10k_PBMC_ATAC_nextgem_Chromium_X_fragments.tsv.gz"
    ),
    annotation = annotation
)

# Requantify query ATAC to match multiome features
requant_multiome_atac <- Signac::FeatureMatrix(
    fragments = Signac::Fragments(atac_pbmc_assay),
    features = Signac::granges(obj_multi[["ATAC"]]),
    cells = Seurat::Cells(atac_pbmc_assay)
)

# Create assay with requantified ATAC data
atac_assay <- Signac::CreateChromatinAssay(
    counts = requant_multiome_atac,
    fragments = file.path(
        directory,
        "inputs",
        "cross_modality_integration",
        "10k_PBMC_ATAC_nextgem_Chromium_X_fragments.tsv.gz"
    ),
    annotation = annotation
)

# Create Seurat object
obj_atac <- Seurat::CreateSeuratObject(
    counts = atac_assay,
    assay = "ATAC")

# Save
# saveRDS(
#     object = obj_atac,
#     file = file.path(
#         directory,
#         "outputs",
#         "04_cross_modality_integration",
#         paste0(format(Sys.Date(), "%b_%d_%Y"),
#         "-pbmc_granulocyte_atac_filtered_feature_bc_raw_seurat")
#     )
# )

obj_atac[["peak.orig"]] <- atac_pbmc_assay

# Apply filtering
obj_atac %<>% subset(
    subset = nCount_ATAC < 7e4 & nCount_ATAC > 2000)
```

# Load reference dataset

Note 4: Unable to finish this script due to memory limitations and `rhdf5` incompatibility. The rest of the script will be keep it **without checking the functionality** from this point until the end.

Note 5: vector limitation size is the main reason that block the load of the `rds` object.

```{r}
ref_rna <- readRDS(
    file.path(
        directory,
        "inputs",
        "cross_modality_integration",
        "pbmc_multimodal_2023.rds"))
```

# Prepocessing and normalization

```{r}
# Multiome RNA by SCTransform
Seurat::DefaultAssay(obj_multi) <- "RNA"
obj_multi %<>% Seurat::SCTransform(verbose = FALSE)

# Multiome ATAC by TF-IDF normalization
Seurat::DefaultAssay(obj_multi) <- "ATAC"
obj_multi %<>% Signac::RunTFIDF()
obj_multi %<>% Signac::FindTopFeatures(min.cutoff = "q0")

# Query ATAC
obj_atac %<>% Signac::RunTFIDF()
```

# Bridge integration

```{r}
# Dimensions: drop first LSI dimension (captures sequencing depth, not biology)
dims_atac <- 2:50
dims_rna <- 1:50

Seurat::DefaultAssay(obj_multi) <- "RNA"
Seurat::DefaultAssay(obj_rna) <- "SCT"

# Build extended reference using bridge dataset
obj_rna_ext <- Seurat::PrepareBridgeReference(
    reference = obj_rna,
    bridge = obj_multi,
    reference.reduction = "spca",
    reference.dims = dims_rna,
    normalization.method = "SCT"
)

# Find anchors across modalities
bridge_anchor <- Seurat::FindBridgeTransferAnchors(
    extended.reference = obj_rna_ext,
    query = obj_atac,
    reduction = "lsiproject",
    dims = dims_atac
)

# Map query onto reference
obj_atac %<>% Seurat::MapQuery(
    anchorset = bridge_anchor,
    reference = obj_rna_ext,
    refdata = list(
        l1 = "celltype.l1",
        l2 = "celltype.l2",
        l3 = "celltype.l3"
    ),
    reduction.model = "wnn.umap"
)
```

# Plotting exploration

```{r}
# Unsupervised ATAC processing for validation
obj_atac %<>%
    Signac::FindTopFeatures(min.cutoff = "q0") %>%
    Signac::RunSVD() %>%
    Seurat::RunUMAP(reduction = "lsi", dims = 2:50)

# Coverage plots at canonical loci
Signac::CoveragePlot(
    obj_atac,
    region = "PAX5",
    group.by = "predicted.l1",
    idents = c("B", "CD4 T", "Mono", "NK"),
    window = 200,
    extend.upstream = -150000
)
```

```{r}
sessionInfo()
```